---
title: "Intro to data processing"
subtitle: ""
author: "Jakub Nowosad <br>nowosad.jakub@gmail.com"
date: "2017-04-24"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library('knitr')
opts_chunk$set(cache=FALSE, warning=FALSE, message=FALSE)
options(scipen = 5)
options(htmltools.dir.version = FALSE)
```

## Dataset

```{r, eval=FALSE}
library('gapminder')
```

or

```{r}
gapminder <- readRDS('data/gapminder.rds')
```

http://www.gapminder.org/data/

http://github.com/jennybc/gapminder

http://www.youtube.com/watch?v=jbkSRLYSojo

---
# Vectors - indexing

- empty index - `[ ]`
- positive index - `[5]`
- zero index - `[0]`
- negative index - `[-1]`
- logical index - `[x>5]`
- name index - `["country"]`

---
## data.frame - indexing
- A data frame is (at least) a two-dimensional object
- There are many ways to subset a data frame. Some of them are:
    - `[]` operator
    - `$` operator
    - or `subset()` function

```{r}
head(gapminder)
```

---
## data.frame - indexing

- `$` operator can be used to retrieve a variable (column) by its name
- The result is a vector

```{r}
gapminder$country
```

---
## data.frame - indexing

- `$` operator can be used to retrieve a variable (column) by its name
- The result is a vector

```{r}
countries <- gapminder$country
countries
```

---
## data.frame - indexing

- `[]` operator can be used to subset rows and columns
- `[]` operator requires two arguments - one for rows (observations) and one for columns (variables) - `[rows, columns]`

```{r}
head(gapminder)
```

---
## data.frame - indexing

- `[]` operator can be used to subset rows and columns
- `[]` operator requires two arguments - one for rows (observations) and one for columns (variables) - `[rows, columns]`

```{r}
gapminder[3, ]
```

---
## data.frame - indexing

- `[]` operator can be used to subset rows and columns
- `[]` operator requires two arguments - one for rows (observations) and one for columns (variables) - `[rows, columns]`

```{r}
gapminder[2, 3]
```

---
## data.frame - indexing

- `[]` operator can be used to subset rows and columns
- `[]` operator requires two arguments - one for rows (observations) and one for columns (variables) - `[rows, columns]`

```{r}
gapminder[-(1:1650), 3]
```

---
## data.frame - indexing

- `[]` operator can be used to subset rows and columns
- `[]` operator requires two arguments - one for rows (observations) and one for columns (variables) - `[rows, columns]`

```{r}
gapminder[c(2,4), -3]
```

---
## data.frame - indexing

- `[]` operator can be used to subset rows and columns
- `[]` operator requires two arguments - one for rows (observations) and one for columns (variables) - `[rows, columns]`

```{r}
gapminder[c(2,4), c("country", "pop")]
```

---
## data.frame - indexing

- Logical operators, such as `==, !=, >, >=, <, <=, &, |` can also be used for indexing

```{r}
gapminder[gapminder$lifeExp>82, c("country", "pop")]
```

```{r}
subset(gapminder, pop>1000000000)
```

---
## data.frame - indexing

- Logical operators, such as `==, !=, >, >=, <, <=, &, |` can also be used for indexing

```{r}
subset(gapminder, pop>1000000000 & country=='India')
```

```{r}
subset(gapminder, pop>1000000000 | continent=='Oceania')
```

---
## data.frame - indexing

- A new object can be created from the results of indexing

```{r}
africa <- subset(gapminder, continent=='Africa')
africa
```

---
## data.frame - indexing

- A new object can be created from the results of indexing

```{r}
year_2007 <- gapminder[gapminder$year==2007, ]
year_2007
```

---
## data.frame - indexing

- object[row,column]
```{r, eval=FALSE}
gapminder[2, 1]
gapminder[1, 3]
gapminder[, 1]
gapminder[1]
gapminder[1, ]
gapminder[0, 3]
gapminder[-2, ]
gapminder[, -1]
gapminder[c(1,5,7), ]
gapminder[, c(1,2)]
gapminder[gapminder$pop<100000, ]
gapminder[gapminder$year==1952, c(4)]
```

---
## data.frame - indexing

```{r, eval=FALSE}
gapminder$lifeExp
```

```{r, echo=FALSE}
head(gapminder$lifeExp)
```

```{r, eval=FALSE}
gapminder$continent
```

```{r, echo=FALSE}
head(gapminder$continent)
```
---
## **dplyr**

```{r}
# install.packages('dplyr')
# devtools::install_github("tidyverse/dplyr")
library('dplyr')
```

- **dplyr** is a package for data exploration and transformation
- Basic functions of the **dplyr** package are `select()`, `arrange()`, `filter()`, `mutate()`, and `summarize()`
- Another important function is `group_by()`
- Additionally, **dplyr** package implements a pipe operator (`%>%`), joins and set operations, and moving window operations

---
## **dplyr** - `select()`

- `select()` function picks columns by its name
- Many additional features - `?dplyr::select()`

```{r, echo=FALSE}
knitr::kable(head(gapminder), 'html')
```

---
## **dplyr** - `select()`

```{r}
gapminder_sel1 <- select(gapminder, country, year, pop)
```

```{r, echo=FALSE}
knitr::kable(gapminder_sel1[1:2, ], 'html')
```

```{r}
gapminder_sel2 <- select(gapminder, -continent)
```

```{r, echo=FALSE}
knitr::kable(gapminder_sel2[1:2, ], 'html')
```

---
## **dplyr** - `arrange()`

- `arrange()` reorder rows in ascending order...

```{r}
gapminder_arr1 <- arrange(gapminder, lifeExp)
```

```{r, echo=FALSE}
knitr::kable(gapminder_arr1[1:5, ], 'html')
```

---
## **dplyr** - `arrange()`

- ...or descending order

```{r}
gapminder_arr1 <- arrange(gapminder, desc(lifeExp))
```

```{r, echo=FALSE}
knitr::kable(gapminder_arr1[1:5, ], 'html')
```

---
## **dplyr** - `filter()`

- `filter()` keeps rows matching given criteria
- Logical operators, such as `==, !=, >, >=, <, <=, &, |` can be used

```{r}
gapminder_fil1 <- filter(gapminder, year == 2007)
```

```{r, echo=FALSE}
knitr::kable(gapminder_fil1[1:2, ], 'html')
```

---
## **dplyr** - `filter()`

```{r}
gapminder_fil2 <- filter(gapminder, year == 2007 & pop <= 999999)
```

```{r, echo=FALSE}
knitr::kable(gapminder_fil2[1:2, ], 'html')
```

```{r}
gapminder_fil2 <- filter(gapminder, continent != 'Asia' |  lifeExp < 30)
```

```{r, echo=FALSE}
knitr::kable(gapminder_fil2[1:2, ], 'html')
```

---
## **dplyr** - `mutate()`

- `mutate()` function creates new variables based on existing variables

```{r}
gapminder_mut1 <- mutate(gapminder, gdp_mil = (gdpPercap * pop) / 1000000)
```

```{r, echo=FALSE}
knitr::kable(gapminder_mut1[1:5, ], 'html')
```

---
## **dplyr** - `summarize()`

- `summarize()` function can be used to reduce variables to values

```{r}
gapminder_sum1 <- summarize(gapminder, mean_le = mean(lifeExp))
```

```{r, echo=FALSE}
gapminder_sum1
```

```{r}
gapminder_sum2 <- summarize(gapminder, 
                            mean_le = mean(lifeExp), 
                            min_le = min(lifeExp), 
                            max_le = max(lifeExp))
```

```{r, echo=FALSE}
gapminder_sum2
```

---
## **dplyr** - `group_by()`

- `group_by()` does not change data. Its role is to create the groups of variables to be used by the previous functions

```{r}
gapminder2007 <- filter(gapminder, year == 2007)
gapminder_grp1 <- group_by(gapminder2007, continent)
```

```{r, echo=FALSE}
head(gapminder_grp1)
```

---
## **dplyr** - `group_by()`

```{r}
gapminder_grp1 <- summarize(gapminder_grp1, mean_lifeExp = mean(lifeExp))
```


```{r, echo=FALSE}
knitr::kable(gapminder_grp1[1:5, ], 'html')
```

---
## **tidyr**

```{r}
# install.packages('tidyr')
library('tidyr')
```

- Many of real world datasets are messy - they are not organized in a way we (or R functions) expect. It is important to use "tidy data" framework to ease the manipulation, modeling, and visualisation of datasets. In "tidy data":
    - Each variable forms a column
    - Each observation forms a row
    - Each type of observational unit forms a table (data frame)
- There are two data formats in "tidy data" - wide format and long format 
- More information about "tidy data" can be found here - http://vita.had.co.nz/papers/tidy-data.html
- The **tidyr** package has two main functions - `gather()` and `spread()`, and several additional functions

---
## **tidyr** - `spread()`

- `spread()` takes two columns (key & value) and spreads them into multiple columns (creates data in a "wide" format)

```{r}
gapminder_sel <- select(gapminder, -continent, -lifeExp, -gdpPercap)
gapminder_spread <- spread(gapminder_sel, country, pop)
```

```{r, echo=FALSE}
knitr::kable(gapminder_spread[c(1:5), 1:10], 'html')
```

---
## **tidyr** - `gather()`

- `gather()` reshapes data from a wide format to a long format

```{r}
gapminder_gather <- gather(gapminder_spread, country, pop, -year)
```

```{r, echo=FALSE}
knitr::kable(gapminder_gather[1:5, ], 'html')
```

---
## `%>%` operator

The usual approach to perform numerous operations in R is either:

- by creating many "subobjects":

```{r}
gapminder_sel <- select(gapminder, country, continent, pop)
gapminder_africa <- filter(gapminder_sel, continent=='Africa')
```

- ...or by nesting:

```{r}
gapminder_africa <- filter(select(gapminder, country, continent, pop), continent=='Africa')
```

```{r, echo=FALSE}
knitr::kable(gapminder_africa[1:5, ], 'html')
```

---
## `%>%` operator

- Nowadays, it could be easier done with the use of the `%>%` ("pipe") operator (**magrittr** or **dplyr** package)
- In this approach, commands are written in a (more) natural order

```{r}
gapminder_africa <- gapminder %>% 
        select(country, continent, pop) %>% 
        filter(continent=='Africa')
```

```{r, echo=FALSE}
knitr::kable(gapminder_africa[1:5, ], 'html')
```

---
## `%>%` operator

```{r}
gapminder_proc <- gapminder %>% 
        filter(continent=='Europe', year==2007) %>% 
        mutate(pop_in_thousands=pop/1000) %>% 
        select(country, gdpPercap, pop_in_thousands) %>% 
        gather(key, value, gdpPercap, pop_in_thousands)
```


```{r, echo=FALSE}
knitr::kable(gapminder_proc[1:5, ], 'html')
```

---
## Resources

- [R for Data Science](http://r4ds.had.co.nz/) - a great book which contains chapters for beginners and for more advanced users
- [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) - an overview of a "tidy data" approach
- [dplyr](http://dplyr.tidyverse.org/) - an official website of the `dplyr` package
- [tidyr](http://tidyr.tidyverse.org/) - an official website of the `tidyr` package
- [Introduction to dplyr](http://stat545.com/block009_dplyr-intro.html) - a clear introdution to the `dplyr` package
- [Tidy data lesson using Lord of the Rings data](https://github.com/jennybc/lotr-tidy#readme) - an introduction to a "tidy data" and the `tidyr` package using Lord of the Rings data
- [Data Wrangling with dplyr and tidyr Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) - a two-pages summary of `dplyr` and `tidyr` functions
- [Cheatsheet for dplyr join functions](http://stat545.com/bit001_dplyr-cheatsheet.html) - examples of `dplyr` join functions
- [A new data processing workflow for R: dplyr, magrittr, tidyr, ggplot2](http://zevross.com/blog/2015/01/13/a-new-data-processing-workflow-for-r-dplyr-magrittr-tidyr-ggplot2/) - a data processing workflow example